---
title: "STATS485 Unit 1 Paper Appendix"
author: "Brendan Bebb"
date: "2023-01-14"
output: pdf_document
---

```{r setup, include=FALSE}
# External Requirements 
library(knitr)
library(dplyr)
library(magrittr)
library(parallel)
library(parallelly)
library(stats)
mobility = read.csv("http://dept.stat.lsa.umich.edu/~bbh/s485/data/mobility1.csv")
americannationsdata = read.csv("https://tinyurl.com/yb65jsq2")
```

### Overview and Goals
The purpose of this appendix is to highlight and construct confidence intervals
for a dataset of commuting zones and their respective citizens. The data contains
a sampling of teenage kids and their current economic status at that time, and then
compares that to their economic status about ten years later. From this information,
a proportion is calculated to show those individuals who move from the lower quintile
at ages 16-18 to the upper quintile of economic status at age 30.

Given the sample size and the desire to capture the true population proportion of
citizens that move from the lowest quintile to the highest quintile, I believe
the Jeffreys' Interval is the best interval to use to find the true population
proportion of individuals with the greatest economic mobility since we are dealing
with such a small p and a large n. Below I will replicate Jeffreys' Interval,
and then modify it based on Brown, Cai, and Disgupta's recommendations to enhance it.

### Computations and Calculations

First, it is necessary to add a column to the dataset to show the number of people
that were sampled who moved from the lower to the upper quintile.

```{r}
mobility$n_upmovers = mobility$n_lowstart * mobility$phat_upmover
```

Next, a Jeffrey's Interval function is implemented to get both the upper and lower
bounds of the intveral.

```{r}
# lower bound
JeffreysInterval_lower = function(n, k, phat) {
  x = phat * n
  alpha = (1 - k)/ 2
  kappa = qnorm(1 - alpha)
  w_t1 = ((kappa * sqrt(4 *  phat * (1 - phat)) / n) + ((kappa^2 - 3) / (6 * n^2))) / (4 * phat * (1 - phat))
  w_t2 = ((0.5 - phat) * (phat * (1-phat) * (kappa^2 + 2) - (1/n))) / (6 * n * (phat * (1-phat))^2)
  w = w_t1 + w_t2
  if (phat == 0) {
    lowerbound = 0
  }
  else if (phat == 1) {
    lowerbound = (x + 0.5) / (n + 1)
  }
  else {
    lowerbound = (x + 0.5) / (n + 1 +(n - x + 0.5) * (exp(2*w) - 1))
  }
  return(lowerbound)
}

# upper bound
JeffreysInterval_upper = function(n, k, phat) {
  x = phat * n
  alpha = (1 - k)/ 2
  kappa = 0 - qnorm(1 - alpha)
  w_t1 = ((kappa * sqrt(4 *  phat * (1 - phat)) / n) + (kappa^2 - 3) / (6 * n^2)) / (4 * phat * (1 - phat))
  w_t2 = ((0.5 - phat) * (phat * (1-phat) * (kappa^2 + 2) - (1/n))) / (6 * n * ((phat * (1-phat))^2))
  w = w_t1 + w_t2
  if (phat == 0) {
    upperbound = (x + 0.5) / (n + 1)
  }
  else if (phat == 1) {
    upperbound = 1
  }
  else {
    upperbound = (x + 0.5) / (n + 1 +(n - x + 0.5) * (exp(2*w) - 1))
  }
  return(upperbound)
}

JeffreysInterval = function(n, k, phat) {
  interval = c()
  interval[1] = JeffreysInterval_lower(n, k, phat)
  interval[2] = JeffreysInterval_upper(n, k, phat)
  return(interval)
}
```

Below are test cases carrying out the Jeffreys' Interval.

```{r}
test1 = JeffreysInterval(30, .95, .1)
test1
test2 = JeffreysInterval(30, .95, 0)
test2
test3 = JeffreysInterval(30, .95, 1)
test3
test4 = JeffreysInterval(30, .75, .41)
test4
test4 = JeffreysInterval(100, .75, .41)
test4
test5 = JeffreysInterval(10, .95, .99)
test5
```

The upper and lower bounds of a 95% Jeffreys' Interval are added to each commuting
zone for easier reference.

```{r}
mobility$lowerbound = 0
mobility$upperbound = 0
for (i in 1:40) {
  mobility[i,9] = JeffreysInterval_lower(mobility[i,6], 0.95, mobility[i,7])
  mobility[i,10] = JeffreysInterval_upper(mobility[i,6], 0.95, mobility[i,7])
}
```

Each regions' total populations are added together from their respective commuting
zones to compute the overall region probability of mobility.

```{r}
sumN_west = 0
lowstart_west = 0

sumN_midwest = 0
lowstart_midwest = 0

sumN_northeast = 0
lowstart_northeast = 0

sumN_south = 0
lowstart_south = 0

top_west = 0 
top_midwest = 0 
top_northeast = 0 
top_south = 0 

for (i in 1:10) {
  sumN_west = sumN_west + mobility[i, 4]
  top_west = top_west + (mobility[i,4] * mobility[i,7])
  lowstart_west = lowstart_west + mobility[i,6]
  
  sumN_midwest = sumN_midwest + mobility[i + 10, 4]
  top_midwest = top_midwest + (mobility[i + 10,4] * mobility[i + 10,7])
  lowstart_midwest = lowstart_midwest + mobility[i+10,6]
  
  sumN_northeast = sumN_northeast + mobility[i + 20, 4]
  top_northeast = top_northeast + (mobility[i + 20,4] * mobility[i + 20,7])
  lowstart_northeast = lowstart_northeast + mobility[i + 20,6]
  
  sumN_south = sumN_south + mobility[i + 30, 4]
  top_south = top_south + (mobility[i + 30,4] * mobility[i + 30,7])
  lowstart_south = lowstart_south + mobility[i + 30,6]
}
```

The probabilities for each region are as follows: 

```{r}
phat_west = top_west / sumN_west
phat_west

phat_midwest = top_midwest / sumN_midwest
phat_midwest

phat_northeast = top_northeast / sumN_northeast
phat_northeast

phat_south = top_south / sumN_south
phat_south
```

The intervals for each region are then calculated for a stronger visualization
of the population proportions.

```{r}
westCI = JeffreysInterval(lowstart_west, 0.95, phat_west)
westCI

midwestCI = JeffreysInterval(lowstart_midwest, 0.95, phat_midwest)
midwestCI

northeastCI = JeffreysInterval(lowstart_northeast, 0.95, phat_northeast)
northeastCI

southCI = JeffreysInterval(lowstart_south, 0.95, phat_south)
southCI
```

Nations are then assigned to each commuting zone based off of Woodward's New
York Times article. These are added in another column to each commuting zone.

```{r}
mobility$nation = NA
for (i in 1:40) {
  for (j in 1:278) {
    if(mobility[i, 1] == americannationsdata[j,1]) {
      if(mobility[i,2] == americannationsdata[j,2]) {
        mobility[i,11] = americannationsdata[j,4]
      }
    }
  }
}


# add in other nations not found in americannations dataset
mobility[2,11] = "Far West"
mobility[3,11] = "Far West"
mobility[4,11] = "El Norte"
mobility[10,11] = "Left Coast"
mobility[31,11] = "New France"
mobility[32,11] = "New France"
mobility[37, 11] = "Greater Appalachia"
mobility[38,11] = "El Norte"
mobility[39,11] = "Greater Appalachia"
```

Below is a preview of the commuting zones in descending order of highest to lowest
mobility probabilities.

```{r}
sorted = mobility[order(-mobility$phat_upmover),]
head(sorted)
```

As done above for the regions, the overall populations for each nation are added 
together to find each nation's probability of mobility.

```{r}
sumN_deepsouth = 0
lowstart_deepsouth = 0
top_deepsouth = 0

sumN_elnorte = 0
lowstart_elnorte = 0
top_elnorte = 0

sumN_farwest = 0
lowstart_farwest = 0
top_farwest = 0

sumN_greatapp = 0
lowstart_greatapp = 0
top_deepsouth = 0

sumN_greatapp = 0
lowstart_greatapp = 0
top_greatapp = 0

sumN_leftcoast = 0
lowstart_leftcoast = 0
top_leftcoast = 0

sumN_midlands = 0
lowstart_midlands = 0
top_midlands = 0

sumN_newfrance = 0
lowstart_newfrance = 0
top_newfrance = 0

sumN_newnet = 0
lowstart_newnet = 0
top_newnet = 0

sumN_yankee = 0
lowstart_yankee = 0
top_yankee = 0



for (i in 1:40) {
  if (mobility[i, 11] == "Deep South") {
    sumN_deepsouth = sumN_deepsouth + mobility[i, 4]
    top_deepsouth = top_deepsouth + (mobility[i,4] * mobility[i,7])
    lowstart_deepsouth = lowstart_deepsouth + mobility[i,6]
  }
  else if (mobility[i, 11] == "El Norte") {
    sumN_elnorte = sumN_elnorte + mobility[i, 4]
    top_elnorte = top_elnorte + (mobility[i,4] * mobility[i,7])
    lowstart_elnorte = lowstart_elnorte + mobility[i,6]
  }
  else if (mobility[i, 11] == "Far West") {
    sumN_farwest = sumN_farwest + mobility[i, 4]
    top_farwest = top_farwest + (mobility[i,4] * mobility[i,7])
    lowstart_farwest = lowstart_farwest + mobility[i,6]
  }
  else if (mobility[i, 11] == "Greater Appalachia") {
    sumN_greatapp = sumN_greatapp + mobility[i, 4]
    top_greatapp = top_greatapp + (mobility[i,4] * mobility[i,7])
    lowstart_greatapp = lowstart_greatapp + mobility[i,6]
  }
  else if (mobility[i, 11] == "Left Coast") {
    sumN_leftcoast = sumN_leftcoast + mobility[i, 4]
    top_leftcoast = top_leftcoast + (mobility[i,4] * mobility[i,7])
    lowstart_leftcoast = lowstart_leftcoast + mobility[i,6]
  }
  else if (mobility[i, 11] == "Midlands") {
    sumN_midlands = sumN_midlands + mobility[i, 4]
    top_midlands = top_midlands + (mobility[i,4] * mobility[i,7])
    lowstart_midlands = lowstart_midlands + mobility[i,6]
  }
  else if (mobility[i, 11] == "New France") {
    sumN_newfrance = sumN_newfrance + mobility[i, 4]
    top_newfrance = top_newfrance + (mobility[i,4] * mobility[i,7])
    lowstart_newfrance = lowstart_newfrance + mobility[i,6]
  }
  else if (mobility[i, 11] == "New Netherland") {
    sumN_newnet = sumN_newnet + mobility[i, 4]
    top_newnet = top_newnet + (mobility[i,4] * mobility[i,7])
    lowstart_newnet = lowstart_newnet + mobility[i,6]
  }
  else {
    sumN_yankee = sumN_yankee + mobility[i, 4]
    top_yankee = top_yankee + (mobility[i,4] * mobility[i,7])
    lowstart_yankee = lowstart_yankee + mobility[i,6]
  }
}
```

The probabilities of mobility for each nation are as follows:

```{r}
phat_ds = top_deepsouth / sumN_deepsouth
phat_ds

phat_en = top_elnorte / sumN_elnorte
phat_en

phat_fw = top_farwest / sumN_farwest
phat_fw

phat_ga = top_greatapp / sumN_greatapp
phat_ga

phat_nn = top_newnet / sumN_newnet
phat_nn

phat_nf = top_newfrance / sumN_newfrance
phat_nf

phat_yank = top_yankee / sumN_yankee
phat_yank

phat_lc = top_leftcoast / sumN_leftcoast
phat_lc

phat_midland = top_midlands / sumN_midlands
phat_midland
```

Below are Jeffreys' intervals for each nation.

```{r}
# confidence intervals for each nation

dsCI = JeffreysInterval(lowstart_deepsouth, 0.95, phat_ds)
dsCI

enCI = JeffreysInterval(lowstart_elnorte, 0.95, phat_en)
enCI

fwCI = JeffreysInterval(lowstart_farwest, 0.95, phat_fw)
fwCI

gaCI = JeffreysInterval(lowstart_greatapp, 0.95, phat_ga)
gaCI

nnCI = JeffreysInterval(lowstart_newnet, 0.95, phat_nn)
nnCI

nfCI = JeffreysInterval(lowstart_newfrance, 0.95, phat_nf)
nfCI

lcCI = JeffreysInterval(lowstart_leftcoast, 0.95, phat_lc)
lcCI

midlandCI = JeffreysInterval(lowstart_midlands, 0.95, phat_midland)
midlandCI

yankCI = JeffreysInterval(lowstart_yankee, 0.95, phat_yank)
yankCI
```

A potential alternative variable worth investigating to see if it has a potential
impact on economic mobility is the population size of each commuting zone and
assigning it to a population size "bracket". The brackets are as follows:

Bracket Group | Population Range
--------------|-----------------
Bracket 1     | 0 - 50,000
Bracket 2     | 50,001 - 100,000
Bracket 3     | 100,001 - 150,000
Bracket 4     | 150,001 - 200,000
Bracket 5     | 200,001+

```{r}
mobility$bracket = NA
for (i in 1:40) {
  if (mobility[i, 4] <= 50000) {
    mobility[i, 12] = 1
  }
  else if (mobility [i, 4] <= 100000) {
    mobility[i, 12] = 2
  }
  else if (mobility [i, 4] <=150000) {
    mobility[i, 12] = 3
  }
  else if (mobility [i, 4] <=200000) {
    mobility[i, 12] = 4
  }
  else {
    mobility[i, 12] = 5
  }
}
```

The probabilities of each of these brackets are then found by adding together the
populations of each bracket, as previously done for the nation and region distribution.

```{r}

sumN1 = 0
top1 = 0
lowstart1 = 0
sumN2 = 0
top2 = 0
lowstart2 = 0
sumN3 = 0
top3 = 0
lowstart3 = 0
sumN4 = 0
top4 = 0
lowstart4 = 0
sumN5 = 0
top5 = 0
lowstart5 = 0

for (i in 1:40) {
  if (mobility[i, 12] == 1) {
    sumN1 = sumN1 + mobility[i, 4]
    top1 = top1 + (mobility[i,4] * mobility[i,7])
    lowstart1 = lowstart1 + mobility[i,6]
  }
  else if (mobility[i, 12] == 2) {
    sumN2 = sumN2 + mobility[i, 4]
    top2 = top2 + (mobility[i,4] * mobility[i,7])
    lowstart2 = lowstart2 + mobility[i,6]
  }
  else if (mobility[i, 12] == 3) {
    sumN3 = sumN3 + mobility[i, 4]
    top3 = top3 + (mobility[i,4] * mobility[i,7])
    lowstart3 = lowstart3 + mobility[i,6]
  }
  else if (mobility[i, 12] == 4) {
    sumN4 = sumN4 + mobility[i, 4]
    top4 = top4 + (mobility[i,4] * mobility[i,7])
    lowstart4 = lowstart4 + mobility[i,6]
  }
  else {
    sumN5 = sumN5 + mobility[i, 4]
    top5 = top5 + (mobility[i,4] * mobility[i,7])
    lowstart5 = lowstart5 + mobility[i,6]
  }
}
```

The probability of economic mobility in each bracket are as follows:

```{r}
phat1 = top1 / sumN1
phat1

phat2 = top2 / sumN2
phat2

phat3 = top3 / sumN3
phat3

phat4 = top4 / sumN4
phat4

phat5 = top5 / sumN5
phat5
```

